Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd

author-running: T. Khan et al.
abstract: |
  Recent advancements in deep learning have significantly enhanced the extraction of detailed ecological information from multispectral (RGBi) aerial imagery, notably the identification and segmentation of individual tree crowns. Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges. Even minor inaccuracies in tree crown segmentation can substantially impact practical applications in forestry management and urban planning. Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy. Addressing these issues necessitates careful selection and design of training data and exploration of task-specific pre-training through Foundation Models. We highlight these validation complexities and recommend for rigorous steps to ensure the reliability and practical utility of large-scale deep learning models in ecological and urban management contexts.
keywords: [Deep Learning, Ecology, Forestry, Remote Sensing.]
---

@@

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).

>>>>

author-running: T. Khan et al.
abstract: |
  Recent advancements in deep learning have significantly enhanced the extraction of detailed ecological information from multispectral (RGBi) aerial imagery, notably the identification and segmentation of individual tree crowns. Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges. Even minor inaccuracies in tree crown segmentation can substantially impact some practical applications in forestry management and urban planning. Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy. Addressing these issues necessitates careful selection and design of training data and exploration of task-specific pre-training through Foundation Models. We highlight these validation complexities and recommend for rigorous steps to ensure the reliability and practical utility of large-scale deep learning models in ecological and urban management contexts.
keywords: [Deep Learning, Ecology, Forestry, Remote Sensing.]
---

@@

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). <!--The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).-->

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
